---
title: "Student Mental Health & Study Habits Analysis"
subtitle: "ETC5513 Assignment 3"
author: "Group 2: Member 1, Member 2, Member 3, Member 4 — ETC5513"
date: "May 2026"

---

# Executive Summary

Student mental health has emerged as a critical concern across university
campuses globally [@hunt2010mental]. This study analyses survey data from
1,200 students to examine how sleep quality, study habits, and counseling
access relate to stress and anxiety outcomes. Students reporting good sleep
quality showed stress levels 3.1 points lower than poor sleepers, and those
with counseling access reported 2.1 points lower anxiety scores. These
findings highlight sleep optimisation and expanded counseling services as
practical, evidence-based interventions for improving student wellbeing
[@walker2017sleep].

# Introduction

Student mental health has become a growing priority for universities
worldwide. Academic pressure, financial stress, and social adjustment
create significant psychological burden for many students. Research
consistently shows that 60-70% of university students experience moderate
to severe stress during their studies [@gaultney2010prevalence]. Poor
mental health is associated with lower academic performance, higher dropout
rates, and long-term health consequences. Daily lifestyle habits including
sleep, exercise, and study patterns play an important role in shaping
mental health outcomes. However, most existing research focuses on general
adult populations rather than student-specific contexts. Understanding which
modifiable factors most strongly predict stress and anxiety in students can
guide targeted campus interventions. This analysis uses cross-sectional
survey data from 1,200 university students to examine these relationships.
Variables include sleep quality, weekly study hours, GPA, anxiety scores,
stress levels, and access to counseling services. Findings from this study
aim to inform practical recommendations for both students and university
administrators.

## Methodology

This analysis examined relationships between study habits, sleep quality,
and mental health outcomes in university students using cross-sectional
survey data [@gaultney2010prevalence]. Variables collected included weekly
study hours, sleep quality (1--10 scale), stress levels (1--10), anxiety
scores (1--10), GPA, and access to counseling services.

Data were loaded using R's tidyverse package and cleaned by removing
records with missing values. Variable names were standardised using the
janitor package [@hunt2010mental]. Categorical variables for sleep quality
(Poor, Fair, Good) and stress levels (Low, Moderate, High) were created
to support group comparisons and visualisation.

Pearson correlations examined bivariate relationships between continuous
variables. Independent samples t-tests compared anxiety levels between
students with and without counseling access. All analyses were conducted
in R (v4.4.0) using tidyverse for data management and ggplot2 for
visualisation.

@tbl-summary presents the sample characteristics including mean study
hours, GPA, stress, anxiety, and sleep quality scores. @fig-pipeline
illustrates the data processing steps applied in this study.
```{r}
#| label: setup-m2
#| include: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
student_data_clean <- read_csv("data/processed/student_data_clean.csv",
                               show_col_types = FALSE)
```

```{r}
#| label: tbl-summary
#| echo: false
#| message: false
#| tbl-cap: "Sample characteristics of university students in the analysis (n = 1,200)."
student_data_clean |>
  summarise(
    `Study Hours (M)` = round(mean(hours_studied_per_week, na.rm = TRUE), 1),
    `GPA (M)`         = round(mean(gpa,          na.rm = TRUE), 2),
    `Stress (M)`      = round(mean(stress_level,  na.rm = TRUE), 1),
    `Anxiety (M)`     = round(mean(anxiety_score, na.rm = TRUE), 1),
    `Sleep (M)`       = round(mean(sleep_quality, na.rm = TRUE), 1),
    `N`               = n()
  ) |>
  kable(booktabs = TRUE) |>
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r}
#| label: fig-pipeline
#| echo: false
#| message: false
#| fig-cap: "Data processing pipeline: steps from raw survey data to the cleaned analytic sample."
#| fig-width: 5.5
#| fig-height: 2.2
library(ggplot2)
steps <- data.frame(
  x     = 1:4,
  label = c("Raw Data\n1,200 obs", "Load\ntidyverse",
            "Clean\njanitor",     "Analyse\n1,200 obs"),
  fill  = c("#d62728", "#ff7f0e", "#1f77b4", "#2ca02c"),
  stringsAsFactors = FALSE
)
arrows_df <- data.frame(
  x = c(1.42, 2.42, 3.42), xend = c(1.58, 2.58, 3.58),
  y = 1, yend = 1
)
ggplot(steps, aes(x = x, y = 1)) +
  geom_tile(aes(fill = fill), width = 0.8, height = 0.6,
            show.legend = FALSE) +
  scale_fill_identity() +
  geom_text(aes(label = label), color = "white",
            size = 2.8, fontface = "bold") +
  geom_segment(data = arrows_df,
               aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.08, "inches")),
               linewidth = 0.7) +
  xlim(0.3, 4.7) + ylim(0.6, 1.4) +
  theme_void()
```

## Results

Analysis of the cleaned dataset (n = 1,200) revealed meaningful relationships
between daily habits and mental health outcomes. Weekly study hours positively
correlated with GPA (r = 0.52, p < 0.001), indicating greater study engagement
predicts better academic performance. Sleep quality showed a strong negative
correlation with stress (r = -0.48, p < 0.001).

Students with counseling access reported significantly lower mean anxiety
(M = 4.2, SD = 2.1) than those without access (M = 6.3, SD = 2.8),
t(1,198) = 9.87, p < 0.001 — a 2.1-point difference.

@fig-study-gpa shows the positive study hours-GPA relationship.
@fig-sleep-stress shows students with good sleep averaging a stress score
of 4.8, compared to 7.9 for poor sleepers — a 3.1-point difference.
@fig-counseling-anxiety confirms the anxiety reduction associated with
counseling access. These findings indicate that sleep quality and counseling
access are the most modifiable predictors of student mental health outcomes.

```{r}
#| label: setup-results
#| include: false
#| message: false
library(tidyverse)
student_data <- read_csv("data/processed/student_data_clean.csv",
                         show_col_types = FALSE) |>
  mutate(sleep_category = factor(sleep_category,
                                 levels = c("Poor", "Fair", "Good")))
```

```{r}
#| label: fig-study-gpa
#| echo: false
#| message: false
#| fig-cap: "Relationship between weekly study hours and GPA (n = 1,200)."
#| fig-width: 5.5
#| fig-height: 3.5
ggplot(student_data, aes(x = hours_studied_per_week, y = gpa)) +
  geom_point(alpha = 0.4, color = "#2E86AB", size = 1.5) +
  geom_smooth(method = "lm", color = "#A23B72", se = TRUE, alpha = 0.2) +
  labs(x = "Hours Studied per Week", y = "GPA (0-10)") +
  theme_minimal(base_size = 11)
```

```{r}
#| label: fig-sleep-stress
#| echo: false
#| message: false
#| fig-cap: "Distribution of stress levels across sleep quality categories."
#| fig-width: 5.5
#| fig-height: 3.5
ggplot(student_data,
       aes(x = sleep_category, y = stress_level, fill = sleep_category)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(
    values = c("Poor" = "#d73027", "Fair" = "#fee090", "Good" = "#1a9850")
  ) +
  labs(x = "Sleep Quality", y = "Stress Level (1-10)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

```{r}
#| label: fig-counseling-anxiety
#| echo: false
#| message: false
#| fig-cap: "Mean anxiety scores by counseling access with 95% confidence intervals."
#| fig-width: 4.5
#| fig-height: 3.5
student_data |>
  group_by(counseling_access) |>
  summarise(
    mean_anxiety = mean(anxiety_score, na.rm = TRUE),
    se           = sd(anxiety_score, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(ci_low  = mean_anxiety - 1.96 * se,
         ci_high = mean_anxiety + 1.96 * se) |>
  ggplot(aes(x = counseling_access, y = mean_anxiety, fill = counseling_access)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                width = 0.2, linewidth = 0.9) +
  scale_fill_manual(values = c("No" = "#d73027", "Yes" = "#1a9850")) +
  labs(x = "Counseling Access", y = "Mean Anxiety (1-10)") +
  theme_minimal(base_size = 11)
```
